Improved Particle Approximations \\ to the Joint Smoothing Distribution \\ Using Markov Chain Monte Carlo

### 2012-02-28 Paper/complete/finaldraft_MCMCsmoother.tex
Added rejection sampling to code. Added references to Douc2009 to paper.

\begin{abstract}
Particle filtering and smoothing algorithms approximate posterior state distributions with a set of samples drawn from those distributions. Conventionally, samples from the joint smoothing distribution are generated by sequentially resampling from the particle filter results. If the number of filtering particles is high, this process is limited by computational complexity. In addition, the support of the smoothing distribution is restricted to the values which appear in the filtering approximation. In this paper, a Markov chain Monte Carlo sampling procedure is used to improve the efficiency of the particle smoother. In addition, an algorithm for approximating the joint smoothing distribution without limited support is presented.
\end{abstract}

@@

The complexity of both the backward-resampling smoother of \cite{Godsill2004} and the two-filter variant of \cite{Briers2010} is $\mathcal{O}(N_F \times N_S \times K)$, where $N_F$ is the number of filter particles, $N_S$ the number of smoother particles, and $K$ the number of time steps. Methods for approximation of such algorithms with lower complexity are discussed in \cite{Klaas2006}. However, computational cost is often prohibitive, requiring either $N_S$ or $N_F$ to be held very low.

@@

@@

The computational complexity of this algorithm is $\mathcal{O}(M \times N_S \times K)$, the same as the MCMC resampling smoother, although it is likely take more time to run because of the additional sampling operation and probability calculations. Because the support of the smoother particles is not restricted to the states which appear in the filter approximation, the error performance is expected to be improved.

@@

The $x_k$ proposal density may be approximated with linear-Gaussian dynamics, e.g. using an extended \cite{Bar-Shalom2002} or unscented \cite{Julier2004} Kalman filter. The weights for the $x_{1:k-1}$ proposal may be approximated using similar methods, but this will often be undesirable. If it is necessary to calculate proposal weights for all of the filter-smoother particles, then the computational advantages of using MCMC are lost. With this in mind, an appropriate choice for the proposal weights might be simply $\tilde{v}_{k-1}^{(j)} = w_{k-1}^{(j)}$.

@@

@@

Tables~\ref{tab:case1_performance} and~\ref{tab:case2_performance} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms in the two noise scenarios, along with running times (excluding the time for filtering) and the mean number of unique smoother particles at each time instant. The trade-off between speed and error is shown in Figs.~\ref{fig:case1_rmse_vs_time} and~\ref{fig:case2_rmse_vs_time}.

\begin{table}[!t]%
\renewcommand{\arraystretch}{1.3}
\caption{Smoother performance measures for noise covariance case 1. All results are averaged over 10 runs, with 500 time steps in each run.}
\label{tab:case1_performance}
\centering
\begin{tabular}{|c||c|c|c|c|}
\hline
Algorithm & \begin{minipage}[c]{1cm} Position RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Velocity RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Running Time (s) \end{minipage} & \begin{minipage}[c]{1cm}  Unique Particles \end{minipage} \\
\hline
FS 						& 0.578 & 0.968 & 0 & 2.25 \\
DBRS					& 0.475 & 0.762 & 122 & 20.3 \\
\hline
MCMC-BRS (1)	& 0.509 & 0.823 & 3 & 13.7 \\
MCMC-BRS (3)	& 0.493 & 0.790 & 7 & 17.4 \\
MCMC-BRS (10)	& 0.482 & 0.768 & 21 & 19.6 \\
MCMC-BRS (30)	& 0.480 & 0.761 & 62 & 20.2 \\
MCMC-BRS (100)& 0.476 & 0.764 & 202 & 20.4 \\
\hline
MCMC-BSS (1)	& 0.473 & 0.758 & 14 & 44.3 \\
MCMC-BSS (3)	& 0.461 & 0.734 & 29 & 69.7 \\
MCMC-BSS (10)	& 0.451 & 0.720 & 82 & 89.8 \\
MCMC-BSS (30)	& 0.444 & 0.713 & 232 & 96.7 \\
MCMC-BSS (100)& 0.443 & 0.709 & 755 & 98.6 \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{case1_smoother_comparison_posRMSE_time.pdf}%
\caption{RMSEs and running times of various smoothing algorithms}%
\label{fig:case1_rmse_vs_time}%
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=0.9\columnwidth]{case2_smoother_comparison_posRMSE_time.pdf}%
\caption{RMSEs and running times of various smoothing algorithms}%
\label{fig:case2_rmse_vs_time}%
\end{figure}

>>>>

\begin{abstract}
Particle filtering and smoothing algorithms approximate posterior state distributions with a set of samples drawn from those distributions. Conventionally, samples from the joint smoothing distribution are generated by sequentially resampling from the particle filter results. If the number of filtering particles is high, this process is limited by computational complexity. In addition, the support of the smoothing distribution is restricted to the values which appear in the filtering approximation. In this paper, a Markov chain Monte Carlo sampling procedure is used to improve the efficiency of the particle smoother, achieving comparable error performance but with a lower execution time. In addition, an algorithm for approximating the joint smoothing distribution without limited support is presented, which achieves simultaneous improvements in both execution time and error. These algorithms also provide a greater degree of flexibility over existing methods, allowing a trade-off between execution time and error, controlled by the length of the Markov chains.
\end{abstract}

@@

The complexity of both the backward-resampling smoother of \cite{Godsill2004} and the two-filter variant of \cite{Briers2010} is $\mathcal{O}(N_F \times N_S \times K)$, where $N_F$ is the number of filter particles, $N_S$ the number of smoother particles, and $K$ the number of time steps. In \cite{Douc2009}, a theoretically linear-cost scheme using rejection sampling is outlined, but in practice this appears to be slower than the standard algorithms. Methods for approximation of such algorithms with lower complexity are discussed in \cite{Klaas2006}. However, computational cost is often still prohibitive, requiring either $N_S$ or $N_F$ to be held very low.

@@

The new scheme targets the same empirical approximation to the smoothing distribution as used in the direct backward-resampling method. However, rather than drawing independent samples from this approximation, MCMC is used, resulting in dependent samples.  In the limit as the number of MH steps, $M$, tends to infinity, the Markov chain will produce an independent sample from the target distribution, as the initial state of the chain will have been forgotten. At this point, the performance of the MCMC method is expected to match that of the direct method in terms of error and particle diversity. At the other extreme, with $M=0$, the algorithm simply outputs particles from the filter-smoother approximation, which is expected to give worse performance. With intermediate values of $M$, the quality of the approximation is expected to fall between those of the filter-smoother and direct backward-resampling smoother.

This new scheme bears a resemblance to the ``linear cost'' simulation scheme of \cite{Douc2009} which uses rejection sampling instead of MCMC. Although theoretically elegant, the latter algorithm has been found to suffer from such high rejection rates as to render it consistently slower than the direct sampling implementation on a practical problem (see section~\ref{sec:simulations}).

@@

The computational complexity of this algorithm is $\mathcal{O}(M \times N_S \times K)$, the same as the MCMC resampling smoother, although it is likely take more time to run because of the additional sampling operation and probability calculations. In contrast with the MCMC resampling method, this smoother targets the exact posterior distribution for the current state at time $t_n$, rather than a particle approximation. Consequently, the performance is expected to be improved relative to the resampling-based methods in terms of both error and particle diversity.

@@

The $x_k$ proposal density may be approximated with linear-Gaussian dynamics, e.g. using an extended \cite{Bar-Shalom2002} or unscented \cite{Julier2004} Kalman filter. The weights for the $x_{1:k-1}$ proposal may be approximated using similar methods, but this will often be undesirable. If it is necessary to calculate proposal weights for all of the filter-smoother particles, then the computational advantages of using MCMC are lost. With this in mind, an appropriate choice for the proposal weights might be simply $\tilde{v}_{k-1}^{(j)} = w_{k-1}^{(j)}$, which avoids calculation of $p(y_{k}, \tilde{x}_{k+1}| x_{1:k-1}^{(j)})$ for all $j$.

@@

The rejection-sampling version of the DBRS described by \cite{Douc2009} was also implemented, but high rejection rates meant that it was impractically slow so it was omitted from further testing.

@@

Tables~\ref{tab:case1_performance} and~\ref{tab:case2_performance} show the root-mean-square errors (RMSEs) in position and velocity obtained by the different algorithms in the two noise scenarios, along with running times (excluding the time for filtering) and the mean number of unique smoother particles at each time instant. The trade-off between speed and error is shown in Fig.~\ref{fig:rmse_vs_time}.

\begin{table}[!t]%
\renewcommand{\arraystretch}{1.3}
\caption{Smoother performance measures for noise covariance case 1. All results are averaged over 10 runs, with 500 time steps in each run.}
\label{tab:case1_performance}
\centering
\begin{tabular}{|c||c|c|c|c|}
\hline
Algorithm & \begin{minipage}[c]{1cm} Position RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Velocity RMSE \end{minipage} & \begin{minipage}[c]{1cm}  Running Time (s) \end{minipage} & \begin{minipage}[c]{1cm}  Unique Particles \end{minipage} \\
\hline
FS 						& 0.578 & 0.968 & 0 & 2.25 \\
DBRS					& 0.475 & 0.762 & 122 & 20.3 \\
\hline
MCMC-BRS (1)    & 0.509 & 0.823 & 3 & 13.7  \\
MCMC-BRS (3)    & 0.493 & 0.790 & 7 & 17.4  \\
MCMC-BRS (10)   & 0.482 & 0.768 & 21 & 19.6 \\
MCMC-BRS (30)   & 0.480 & 0.761 & 62 & 20.2 \\
MCMC-BRS (100)  & 0.476 & 0.764 & 202 & 20.4  \\
\hline
MCMC-BSS (1)	& 0.473 & 0.758 & 14 & 44.3 \\
MCMC-BSS (3)	& 0.461 & 0.734 & 29 & 69.7 \\
MCMC-BSS (10)	& 0.451 & 0.720 & 82 & 89.8 \\
MCMC-BSS (30)	& 0.444 & 0.713 & 232 & 96.7 \\
MCMC-BSS (100)& 0.443 & 0.709 & 755 & 98.6 \\
\hline
\end{tabular}
\end{table}

@@

\begin{figure}[!t]
\centering
\subfloat[Case 1]{\includegraphics[width=0.9\columnwidth]{case1_smoother_comparison_posRMSE_time.pdf}}
\\
\subfloat[Case 2]{\includegraphics[width=0.9\columnwidth]{case2_smoother_comparison_posRMSE_time.pdf}}
\caption{Position RMSEs and running times of various smoothing algorithms with different noise covariance cases.}%
\label{fig:rmse_vs_time}%
\end{figure}
